---
title: "Getting started with RNAflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with RNAflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
have_deseq2 <- requireNamespace("DESeq2", quietly = TRUE)
```

RNAflow is an end-to-end bulk RNA-seq analysis platform, usable either as an
interactive Shiny app (`run_app()`) or as a set of pure, scriptable functions.
This vignette is a **reproducible reference analysis**: the core steps below run
on a bundled, published dataset so you can reproduce a real result end to end.

```{r setup}
library(RNAflow)
```

## The data

The package bundles two real, published human datasets. Here we use
**airway** (Himes *et al.* 2014): airway smooth-muscle cells treated with
dexamethasone vs. control across four cell lines.

```{r load}
counts <- read_counts(system.file("extdata", "demo_airway_counts.csv",
                                  package = "RNAflow"))
meta   <- read_metadata(system.file("extdata", "demo_airway_metadata.csv",
                                    package = "RNAflow"),
                        counts_samples = colnames(counts))
dim(counts)
head(meta)
```

`read_counts()` and `read_metadata()` run strict validation and fail fast with
an explicit message on malformed input (negative values, duplicate gene IDs,
missing rownames, sample mismatch, …).

## Differential expression

We model expression as a function of `condition`, adjusting for the paired
`cell` line, and contrast dexamethasone (`Dex`) against `Control`.

```{r de, eval = have_deseq2}
res <- run_deseq2(
  counts, meta,
  design   = ~ cell + condition,
  contrast = c("condition", "Dex", "Control"),
  shrink   = TRUE            # apeglm LFC shrinkage, Wald stat preserved
)
head(res[order(res$padj), ])
sum(res$padj < 0.05 & abs(res$log2FoldChange) > 1, na.rm = TRUE)  # sig genes
```

## Visualize

```{r volcano, eval = have_deseq2}
fig_volcano(res, lfc_thr = 1, padj_thr = 0.05, n_label = 15,
            mode = "exploration")
```

Every figure function takes a `mode = c("exploration", "publication")` argument;
publication mode uses 8-pt Helvetica, strict axes, and no grid, ready for a
figure panel. Interactive versions (`fig_volcano_interactive()`,
`fig_pca()`, `fig_umap()`, …) return plotly widgets for exploration.

For a heatmap and a PCA / UMAP sample overview you also need a normalized
matrix:

```{r norm, eval = FALSE}
vst <- normalize_counts(counts, meta, method = "vst")
fig_heatmap(vst, res = res, metadata = meta, n_genes = 40)
fig_pca(vst, meta, n_top = 500, color_by = "condition")
fig_umap(vst, meta, color_by = "condition")            # non-linear embedding
```

## Downstream analyses

The heavier steps depend on optional (Bioconductor) packages and network
resources, so they are shown here rather than executed. All work from the same
`res` / `vst` objects and take an `organism` of `"human"`, `"mouse"`, or
`"rat"`.

```{r downstream, eval = FALSE}
# Functional enrichment: GSEA (whole ranked list) and ORA (significant genes)
gene_sets <- get_gene_sets("human", collection = "H")     # MSigDB Hallmark
gsea <- run_gsea(res, gene_sets, rank_by = "stat")
fig_enrich_dot(gsea)

sig <- contrast_sig_genes(res, padj_thr = 0.05, lfc_thr = 1)
ora <- run_ora(sig, "human", db = "GO", ont = "BP", universe = res$gene)
fig_enrich_bar(ora)

# Co-expression network (WGCNA)
datExpr <- wgcna_datexpr(vst, n_genes = 3000)
wg <- run_wgcna(datExpr, power = wgcna_pick_power(datExpr)$suggested)
fig_module_trait(module_trait_cor(wg$MEs, build_traits(meta, rownames(datExpr))))

# Regulator / pathway activity (decoupleR) and per-sample signatures (GSVA)
fig_activity_bar(run_activity(res, "human", what = "pathway"))
fig_gsva_heatmap(run_gsva(vst, gene_sets), meta)
```

## Sessions and reports

Bundle the whole analysis state into a single `.rnaflow.rds` file, and export a
reproducible R script and a self-contained HTML report (no pandoc/Quarto
required):

```{r report, eval = FALSE}
p <- empty_project("airway_study")
p$counts     <- counts
p$metadata   <- meta
p$organism   <- "human"
p$contrasts  <- contrast_store_upsert(list(), "Dex vs Control", res)
save_project(p, "airway_study.rnaflow.rds")

cat(generate_r_script(p), file = "analysis.R")   # runnable Methods script
build_report_html(p, "report.html")              # one-file HTML report
```

## Learn more

Everything above is also available point-and-click in the app:

```{r app, eval = FALSE}
run_app()
```

See the [package website](https://KmBioChemo.github.io/RNAflow/) for the full
function reference and the README for the feature overview.
