---
title: "Getting started with RNAflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with RNAflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

RNAflow is an end-to-end bulk RNA-seq analysis platform. This vignette walks through the typical workflow using a small simulated dataset.

## Launching the app

```{r eval = FALSE}
library(RNAflow)
run_app()
```

The app opens in your browser. The Data tab is the entry point — upload a counts matrix and sample metadata, then run DESeq2.

## Programmatic workflow

You can also use RNAflow's functions in a script.

### Step 1 — Read and validate inputs

```{r eval = FALSE}
counts <- read_counts("path/to/counts.csv")
meta   <- read_metadata("path/to/metadata.csv",
                       counts_samples = colnames(counts))
```

`read_counts()` and `read_metadata()` both run strict validation. You get a clear error message if anything is wrong (negative values, duplicate gene IDs, missing rownames, sample mismatch, etc.).

### Step 2 — Differential expression

```{r eval = FALSE}
res <- run_deseq2(
  counts, meta,
  design   = ~ condition,
  contrast = c("condition", "Treatment", "Control"),
  shrink   = TRUE       # apeglm LFC shrinkage
)
head(res)
```

### Step 3 — Visualize

```{r eval = FALSE}
# Static, publication-ready
p_pub <- fig_volcano(res, lfc_thr = 1, padj_thr = 0.05,
                     n_label = 30, mode = "publication")
save_ggplot(p_pub, "volcano.pdf", "pdf", w = 5, h = 4)

# Interactive (for exploration in a notebook)
fig_volcano_interactive(res)
```

### Step 4 — Heatmap and PCA

For heatmap and PCA you also need a normalized counts matrix:

```{r eval = FALSE}
vst <- normalize_counts(counts, meta, method = "vst")

ph <- fig_heatmap(
  counts_mat = vst, res = res, metadata = meta,
  n_genes = 40, padj_thr = 0.05, lfc_thr = 1,
  palette_name = "RdBu", direction_annotation = TRUE
)
save_pheatmap(ph, "heatmap.pdf", "pdf", w = 6, h = 7)

fig_pca(vst, meta, n_top = 500, color_by = "condition")
```

## Sessions

Save your full analysis state to a `.rnaflow.rds` file:

```{r eval = FALSE}
p <- empty_project("mrl_lpr_hippocampus")
p$counts     <- counts
p$metadata   <- meta
p$de_results <- res
p$organism   <- "mouse"
save_project(p, "mrl_lpr.rnaflow.rds")

# Later...
p <- load_project("mrl_lpr.rnaflow.rds")
```

## Next steps

Phases 3–5 will add GSEA, WGCNA, and a one-click HTML report. See the README roadmap.
